---
title: "leaf_flower_pollen"
author: "Yi"
date: "2023-07-03"
output: html_document
---

```{r setup, echo=FALSE, message=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(see)
```

# leaf flower start date realtion in NPN 

## correlation

We want to check whether some species have fixed lag: y = x+b. Here we can see that maple, birch, pine, oak and elm shows the pattern. Except Elm, other genus shows a y = x pattern which means we may use leaf and flower interchangeably. But let's check the lag between lower and leaf first to see whether the lag is almost zero.

```{r rlm, message=FALSE, warning=FALSE, fig.width=8*1.618 , fig.height=8, cache=TRUE}
source("scripts/plot_rlm_combined_8genuses.R") # after libaray mass, dplyr::select is masked
combined_plot
#rm(list = ls())
```

Fig.1 Leaf and flowering day relationship of pollen allergy related genus. Each dot represents a individual in a year. Each color represent a species. The wide black line represents the total dots robust linear regression. The thin black line is the 1:1 line.


## lag distribution

```{r lag, message=FALSE, warning=FALSE, fig.width=8 , fig.height=8*1.618, cache=TRUE}
source("scripts/plot_quercus_lag_violin.R")
print(lags_violin)
rm(list = ls())
```
Fig.2 Lag of flowering after leafing of Oak. A) Lag distribution by species. B) Zoom in to [-20 ~ 40] range. C) Total lag distribution of oak, average is 15 days.

# using npn to predict pollen

## let's see the result if we supppose leaf and flower is the same and compare their ability to predict pollen
```{r station_lfp, message=FALSE, warning=FALSE, fig.width=8*1.618 , fig.height=8, cache=TRUE}
figures <- read_rds("/nfs/turbo/seas-zhukai/phenology/phenology_leaf_flower_lag/RMSE_Smooth_Quercus_leafflowerpollen.rds")
print(figures[24])
rm(list = ls())
```
Fig.3 Flowering and leafing initiation observations from npn and pollen count from nab.

We can see that: 1) we have more leafing observations than flowering. 2) leafing is a little bit ahead of flowering, adding a lag to leafing may make leaf fits better.

Here are the statistics:

```{r RMSE_sta, message=FALSE, warning=FALSE, fig.width=8*1.618 , fig.height=8,cache=TRUE} 
source("scripts/plot_RMSE_compare.R")
print(RMSE_violin)
rm(list = ls())
```


