---
title: "leafing_flowering_lag"
output: html_document
date: "2023-05-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# download data
```{r}
# source("scripts/download_npn.R")
```

# select data
## datasource: Nature's notebook similar to neon (powerpoint)
## phenophase: 2 & 7 has the most observations (taxa). However 1 & 7 has grass (taxa_1). Both use linear regression

# check distribution of data

# correlation beween flowering and leafing
## linear regression
## robust linear regression
taxa_rlm
## circular-circular statistics
Visual for cc_statistics
1.each genus a pdf
2.combined figure for one species in each genus
```{r}
source("scripts/plot_cc_stat_1speciespergenus.R")
combined_plot
rm(list = ls())
```

# lag and gddlag between flower and leaf
boxplot
histogram
```{r}
source("scripts/plot_histogram_of_lag.R")
```

statistics tibble
delete outliers
what statistics should we use to report the lag (average+std), is the outstanding zero value be a problem?
```{r}
source("scripts/table_lag_statistics.R")
rm(list = ls())
```

```{r}
source("scripts/plot_avail_individual_yr.R")

```



### lag with climate



