---
title: 'explore herbarium data'
embed-resources: true
author: "Yi Liu"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
format:
  html:
    code-folding: show
    highlight: textmate
    number-sections: true
    theme: flatly
    toc: TRUE
    toc-depth: 4
    toc-float:
      collapsed: false
      smooth-scroll: true
---

# get herbarium data

```{r}
library(dplyr)
library(ggplot2)
raw <- read.csv("/Users/yia/Downloads/combined_occ_img_downloaded.csv") %>% 
  dplyr::select(day, month, year, startDayOfYear, coordinateUncertaintyInMeters, decimalLongitude, decimalLatitude, filename_image, species, genus,family)

phenology <- read.csv("/Users/yia/Downloads/phenology.csv") %>% 
  mutate(filename_image = gsub(".txt", "", file_name))

joint_data <- left_join(phenology,raw, by = "filename_image")

```

## generate flowering time
```{r}
summary(joint_data)

joint_data_flower <- joint_data %>% #116805
  mutate(isflower = if_else(flower_one > 0 & flower_many > 0, 1, 0 )) %>% 
  filter(!is.na(startDayOfYear)) %>% # 113253
  filter(isflower==1) %>%  # 21100
  filter(year>=1895) %>%  # 21100
  group_by(species) %>%
  filter(n()>29) %>% # 18849 enough for linear model
  group_by(family,genus) %>%
  ungroup()

```
## clear herbarium data for repeat file and repeat file with different phenology
```{r}
for_climate <- joint_data_flower %>%
  dplyr::select(decimalLongitude, decimalLatitude, startDayOfYear, year, species, genus, family) %>%
  rename(lon = decimalLongitude, lat = decimalLatitude, doy = startDayOfYear) %>% 
  distinct()

# two repeated component:
# 1. completely the same ~300
# 2. same specimen (different name) with different phenology (as long as the flower is consistent, we will keep them) ~2000

```

# check data composition
```{r}
library(data.tree)
# Example data
species_data <- for_climate %>%
  group_by(species, genus, family) %>%
  summarise(specimen_count = n())

# Aggregate data at genus level
genus_data <- species_data %>%
  group_by(family, genus) %>%
  summarise(specimen_count = sum(specimen_count))

# Aggregate data at family level
family_data <- species_data %>%
  group_by(family) %>%
  summarise(specimen_count = sum(specimen_count))

# Total specimen count
total_specimen_count <- sum(species_data$specimen_count)

# Create hierarchical data
hierarchical_data <- species_data %>%
  mutate(level = "species") %>%
  bind_rows(genus_data %>% mutate(species = "", level = "genus")) %>%
  bind_rows(family_data %>% mutate(genus = "", species = "", level = "family")) %>%
  ungroup() %>%
  arrange(family, genus, species)

# Create a data tree structure
tree_data <- hierarchical_data %>%
  mutate(pathString = paste("Total", family, genus, species, sep = "/")) %>%
  select(pathString, specimen_count)

tree <- as.Node(tree_data)
tree$specimen_count <- total_specimen_count

# Print the tree structure to console
print(tree, "specimen_count")

tree_text <- capture.output(print(tree, "specimen_count", pruneMethod = NULL))
writeLines(tree_text, "tree_structure.txt")
```


# extract the climate normality

```{r}
# this part should be able to be speed up by using parellel computing by group them by year and extract all loaction in that year by just reading in the raster file once, and we can assign different cores to work on different years.

library(raster)


# extract the climate normality
complete_period_raster <- raster("../data/prism/complete_period_springmean.tif")

joint_data_flower_normality <- for_climate %>%
  dplyr::select(lat, lon) %>%
  distinct() %>%
  mutate(complete_period_temp = extract(complete_period_raster, cbind(lon, lat)))

# extract the climate anormality
# Initialize an empty data frame to store the results
joint_data_flower_anormality <- data.frame()

# Loop through the specified years
for (fo_year in 1895:2023) {
  # Load the yearly raster file
  yearly_raster <- raster(paste0("../data/prism/", fo_year, "_springmean.tif"))
  
  # Process the joint_data_flower for the current year
  yearly_data <- for_climate %>%
    dplyr::select(year, lat, lon) %>%
    distinct() %>%
    filter(year == fo_year) %>%
    mutate(yearly_temp = extract(yearly_raster, cbind(lon, lat)))
  
  # Append the yearly data to the cumulative data frame
  joint_data_flower_anormality <- rbind(joint_data_flower_anormality, yearly_data)
}

# Combine the normality and anormality data
temperature_data <- joint_data_flower_normality %>%
  right_join(joint_data_flower_anormality, by = c("lat", "lon")) %>%
  rename(norm = complete_period_temp, yeart = yearly_temp) %>%
  mutate(anom = yeart - norm) %>%
  right_join(for_climate, by = c("lat", "lon", "year")) %>%
  filter(!is.na(anom)) %>% #17587->17490
  group_by(species) %>%
  filter(n()>29) %>%   #  enough for linear model 17490->17213
  ungroup()

write.csv(temperature_data, "../data/herb_temperature_data.csv")
```

# visionally check the data
## quercus as an example
```{r}
ggplot(temperature_data %>% filter(genus=="Quercus"), aes(x = yeart, y = doy)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~species) +
  theme(legend.position = "none") 

```

### test on one species: 
#### try to remove outlier using multi-normal distribution
```{r}
library(mvoutlier)

species_data <- temperature_data %>% 
  filter(species=="Quercus wislizeni") 
  
result <- aq.plot(species_data%>% 
  dplyr::select(yeart, doy) %>% 
  as.matrix())
data_clean <- species_data[!result$outliers, ]
```
#### fit and plot the model
```{r}
model <- lm(doy ~ anom+norm,  data_clean)
summary(model)

# calculate condidence interval
model_summary <- broom::tidy(model, conf.int = TRUE)

# Extract coefficients and confidence intervals for anom and norm
coef_anom <- model_summary %>% filter(term == "anom")
coef_norm <- model_summary %>% filter(term == "norm")

# Calculate R-squared
r_squared <- summary(model)$r.squared

# plot the model
# Calculate means of anom, norm, and doy
mean_yeart <- mean(data_clean$yeart, na.rm = TRUE)
mean_doy <- mean(data_clean$doy, na.rm = TRUE)

# Calculate the adjusted intercepts for the center of the figure
intercept_anom <- mean_doy - coef(model)["anom"] * mean_yeart

ggplot(data_clean, aes(x = yeart, y = doy)) +
  geom_point(alpha = 0.5) +
  geom_abline(aes(intercept = intercept_anom, slope = coef(model)["anom"], color = "Temporal")) +
  geom_abline(aes(intercept = coef(model)[1], slope = coef(model)["norm"], color = "Spatial")) +
  labs(title = "Linear Model for Quercus wislizeni", x = "Year", y = "Day of Year") +
  annotate("text", x = min(data_clean$yeart), y = min(data_clean$doy), 
           label = sprintf("Temporal CI: [%0.2f, %0.2f]\nSpatial CI: [%0.2f, %0.2f]\nR² = %0.2f", 
                           coef_anom$conf.low, coef_anom$conf.high, 
                           coef_norm$conf.low, coef_norm$conf.high, 
                           r_squared),
           hjust = 0, vjust = 0, size = 5, color = "black")
```


```{r}
paired_data <- data.frame(
  x_black = data_clean$yeart,
  y_black = data_clean$doy,
  x_red = data_clean$norm,
  y_red = data_clean$doy
)

# Plot the black points
ggplot(paired_data) +
  geom_point(aes(x = x_black, y = y_black, color = "black")) +
  geom_point(aes(x = x_red, y = y_red, color = "red"), alpha = 0.2) +
  geom_segment(aes(x = x_black, y = y_black, xend = x_red, yend = y_red, color = "red"), alpha = 0.2) +
  xlab("spring_ave_temperature") +
  ylab("doy") +
  scale_color_manual(name = "Legend", 
                     values = c("black" = "black", "red" = "red"),
                     labels = c("Observation", "Normality")) +
  geom_abline(intercept = intercept_anom, slope = coef(model)["anom"], color = "blue") +
  geom_abline(intercept = coef(model)[1], slope = coef(model)["norm"], color = "red") 
```

### apply this to all species
```{r}
# Define a function to clean the data for each species
clean_species_data <- function(df) {
  data <- df %>%
    dplyr::select(yeart, doy) %>%
    as.matrix()
  
  result <- aq.plot(data)
  
  # Filter out the outliers
  data_clean <- df[!result$outliers, ]
  return(data_clean)
}

# Apply the cleaning function to each species and bind the results
quercus_clean <- temperature_data %>%
  filter(genus == "Quercus") %>%
  group_by(species) %>%
  group_split() %>%
  purrr::map_dfr(clean_species_data) %>% 
  ungroup()

# Plot the cleaned data
ggplot() +
  geom_point(data = quercus_clean, aes(x = yeart, y = doy), alpha = 0.5) +
  facet_wrap(~species) +
  labs(title = "Cleaned Data for Quercus Species")
```


### fit the model by species and generate a table conclude the confidence interval of the two slpoe for each species
```{r}
source("function_visionalize_summmary_MLmodel.R")
```

```{r}
# plot a figure to show the comparasion of anom slope and norm lope with confidence interval and model fitting
ggplot(summary_results, aes(x = anom_estimate, y = norm_estimate)) +
  geom_errorbar(aes(ymin = norm_conf_low, ymax = norm_conf_high, alpha = r_squared), width = 0) +
  geom_errorbarh(aes(xmin = anom_conf_low, xmax = anom_conf_high, alpha = r_squared), height = 0) +
  geom_point(aes(alpha = r_squared), size = 3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +  # Add 1:1 line
  scale_alpha_continuous(range = c(0.2, 1)) +  # Adjust alpha range for better visualization
  labs(
    title = "Slope Estimates for Anom and Norm by Species",
    x = "Anom Slope Estimate",
    y = "Norm Slope Estimate",
    alpha = "R^2"
  ) +
  theme_minimal() +
  theme(legend.position = "right")


```

kick the outlier
```{r}
  result <- aq.plot(summary_results %>% dplyr::select(anom_estimate, norm_estimate) %>% as.matrix())
  summary_results_clean <- summary_results[!result$outliers, ]
  
  ggplot(summary_results_clean, aes(x = norm_estimate, y = anom_estimate )) +
  geom_errorbarh(aes(xmin = norm_conf_low, xmax = norm_conf_high, alpha = r_squared), width = 0) +
  geom_errorbar(aes(ymin = anom_conf_low, ymax = anom_conf_high, alpha = r_squared), height = 0) +
  geom_point(aes(alpha = r_squared), size = 3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +  # Add 1:1 line
  scale_alpha_continuous(range = c(0.2, 1)) +  # Adjust alpha range for better visualization
  labs(
    # title = "Slope Estimates for Anom and Norm by Species",
    x = "Temporal sensitivity",
    y = "Spatial sensitivity",
    alpha = "R^2"
  ) +
  theme_minimal() +
  theme(legend.position = "right")
  
```

```{r}
summary_results_clean %>% 
#filter for the species that the anom slope and norm slope confidence interval do not overlap
  mutate(anom_conf_low = as.numeric(anom_conf_low),
         anom_conf_high = as.numeric(anom_conf_high),
         norm_conf_low = as.numeric(norm_conf_low),
         norm_conf_high = as.numeric(norm_conf_high)) %>%
  filter(anom_conf_low > norm_conf_high | anom_conf_high < norm_conf_low) %>%
  arrange(desc(r_squared))
```

```{r}
```

```{r}
```

