---
title: 'explore herbarium data'
embed-resources: true
author: "Yi Liu"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
format:
  html:
    code-folding: show
    highlight: textmate
    number-sections: true
    theme: flatly
    toc: TRUE
    toc-depth: 4
    toc-float:
      collapsed: false
      smooth-scroll: true
---

# get herbarium data

```{r}
pheno_all <- read.csv("/Users/yia/Downloads/phenology.csv")

raw_all <- "/Users/yia/Downloads/0011964-240506114902167 (2)/verbatim.txt"

# Read data from a tab-delimited file
data <- read.csv("/Users/yia/Downloads/0011964-240506114902167 (2)/occurrence.csv") 
data_test <- read.delim("/Users/yia/Downloads/0011964-240506114902167 (2)/verbatim.txt", header = TRUE)
data_sm <- data%>% 
  select(day, month, year, startDayOfYear, coordinateUncertaintyInMeters, decimalLongitude, decimalLatitude, filename_image)
```


```{r}
library(dplyr)
library(ggplot2)
raw <- read.csv("/Users/yia/Downloads/Phenology_Acer_saccharum/LM2_output/combined_occ_img_downloaded.csv") %>% 
  select(day, month, year, startDayOfYear, coordinateUncertaintyInMeters, decimalLongitude, decimalLatitude, filename_image)

phenology <- read.csv("/Users/yia/Downloads/Phenology_Acer_saccharum/LM2_output/Phenology/phenology.csv") %>% 
  mutate(filename_image = gsub(".txt", "", file_name))

joint_data <- left_join(phenology,raw, by = "filename_image")
```

# how to decide first leafing day

## method 1: logistic regression (find the classification boundary based on day of year between yes and no)

```{r}
joint_data <- joint_data %>%
  mutate(decade = 10 * floor(year / 10))

# Filter and group the data by decade
joint_data %>%
  filter(!is.na(startDayOfYear)) %>%
  ggplot() +
  geom_point(aes(x = startDayOfYear, y = has_leaves), color = "green", alpha = 0.1) +
  facet_wrap(~ decade, scales = "free_x") +  # Splitting the plot into panels based on decades
  labs(x = "Start Day of Year", y = "Value")  # Adding axis labels
```

```{r}
joint_data %>%
  filter(!is.na(startDayOfYear)) %>%
  ggplot() +
  geom_point(aes(x = startDayOfYear, y = is_fertile), color = "red", alpha = 0.1) +
  facet_wrap(~ decade, scales = "free_x") +  # Splitting the plot into panels based on decades
  labs(x = "Start Day of Year", y = "Value")  # Adding axis labels
```

## method 2: use flower only

```{r}
joint_data_flower <- joint_data %>% 
  mutate(isflower = if_else(flower_one > 0 & flower_many > 0, 1, 0 ))

joint_data_flower %>%
  filter(!is.na(startDayOfYear)) %>%
  ggplot() +
  geom_point(aes(x = startDayOfYear, y = isflower), color = "red", alpha = 0.1) +
  facet_wrap(~ decade, scales = "free_x") +  # Splitting the plot into panels based on decades
  labs(x = "Start Day of Year", y = "Value")  
```

# get climate data for herbarium

## check spatial temporal coverage of the data

```{r}
for_climate <- joint_data_flower %>% 
  filter(isflower==1) %>%  # 254
  filter(year>=1895) # 188

coordinate <- unique(for_climate[, c("decimalLongitude", "decimalLatitude")])
# time distribution
hist(for_climate$year)

plot(for_climate$decimalLongitude,for_climate$decimalLatitude)
```

## download prism

```{r}
# library(prism)
# 
# prism_set_dl_dir("../data/prism")
# 
# get_prism_monthlys(type = "tmean", year = 1895:2023, mon = 3:5, keepZip = FALSE)
# # 
# library(raster)
# library(prism)
# prism_set_dl_dir("data/prism")
# 
# results <- data.frame(year = integer(), longitude = numeric(), latitude = numeric(), spring_avg_temp = numeric())
# 
# # this code can only be run in r script
# # Process each year
# for (focal_year in 1895:2022) {
#   # Load the raster files for March, April, and May
#   tmean_rast_yr_mo <- prism_archive_subset(temp_period = "monthly", type = "tmean", years = focal_year, mon = 3:5)
#   tmean_rast2_yr_mo <- pd_stack(tmean_rast_yr_mo)
#   r_mean <- raster::calc(tmean_rast2_yr_mo, mean) # 1:4 mean for focal_yr
#   #raster::plot(r_mean)
# 
#   tmean_data <- unlist(raster::extract(x = r_mean, 
#                                        y = coordinate)) 
#   
#   results <- rbind(results, data.frame(year = focal_year,
#                                      longitude = coordinate$decimalLongitude,
#                                      latitude = coordinate$decimalLatitude,
#                                      spring_avg_temp = tmean_data))
# }
# 
# library(readr)
# 
# write_rds(results, file = "/Users/yia/Documents/phenology-npn-flower-leaf-lag/data/prism_site_allyear.rds")
```

## connect pheno with climate

```{r}
library(readr)
results <- read_rds("data/prism_site_allyear.rds")
norm_data <- results %>%
  group_by(longitude, latitude) %>%
  summarize(norm = mean(spring_avg_temp, na.rm = TRUE)) %>%    # Calculate group means
  right_join(results, by = c("longitude", "latitude")) %>%    # Join summarized data back to original
  mutate(anom = spring_avg_temp - norm)                       # Calculate anomalies

pheno_with_climate <- for_climate %>%
  rename(longitude = decimalLongitude, latitude = decimalLatitude, doy = startDayOfYear) %>%
  left_join(norm_data, by = c("longitude", "latitude", "year")) %>% 
  filter(!is.na(doy))
```

## visionally check the data

```{r}
# remove doy after 200, and out of us points
pheno_with_climate <- pheno_with_climate %>%
  filter(doy < 200) %>%
  filter(!is.na(norm)) %>%
  distinct(longitude, latitude, year, .keep_all = TRUE) # delete same year same location

paired_data <- data.frame(
  x_black = pheno_with_climate$spring_avg_temp,
  y_black = pheno_with_climate$doy,
  x_red = pheno_with_climate$norm,
  y_red = pheno_with_climate$doy
)

# Plot the black points
ggplot(paired_data) +
  geom_point(aes(x = x_black, y = y_black, color = "black")) +
  geom_point(aes(x = x_red, y = y_red, color = "red"), alpha = 0.2) +
  geom_segment(aes(x = x_black, y = y_black, xend = x_red, yend = y_red, color = "red"), alpha = 0.2) +
  xlab("spring_ave_temperature") +
  ylab("doy") +
  scale_color_manual(name = "Legend", 
                     values = c("black" = "black", "red" = "red"),
                     labels = c("Observation", "Normality"))
```

## fit the multiple regression

```{r}
model1 <- lm(doy~norm+anom,pheno_with_climate)
summary(model1)
confint(model1)

# Creating a new data frame for predictions
norm_range <- seq(min(pheno_with_climate$norm), max(pheno_with_climate$norm), length.out = 100)
anom_range <- seq(min(pheno_with_climate$anom), max(pheno_with_climate$anom), length.out = 100)
grid <- expand.grid(norm = norm_range, anom = anom_range)

# Predicting 'doy' using the model
grid$doy_pred <- predict(model1, newdata = grid)

# Creating the base plot
p <- ggplot() +
  geom_tile(data = grid, aes(x = norm, y = anom, fill = doy_pred)) +
  geom_contour(data = grid, aes(x = norm, y = anom, z = doy_pred), bins = 10, color = "white") +
  labs(title = "Predicted Day of Year from norm and anom", x = "Norm", y = "Anom") +
  theme_minimal()

# Add actual data points using the original data, with colorful fill and white stroke
p <- p + geom_point(data = pheno_with_climate, aes(x = norm, y = anom, fill = doy), color = "white", size = 2, shape = 21, stroke = 0.5)

# Use the same color scale for both the predicted and actual doy
low_color <- "blue"
high_color <- "red"
p <- p + scale_fill_gradient(name = "Day of Year", low = low_color, high = high_color)

# Adjust the aspect ratio so that one unit on the x-axis is the same as one on the y-axis
p <- p + coord_fixed(ratio = 1)

# Combine the legends into one
p <- p + guides(fill = guide_colorbar(title = "Day of Year"), color = "none")

# Render the plot
print(p)
```

# check long term trend

```{r}
library(scales)

ggplot(pheno_with_climate) +
  # geom_point(aes(x = year, y = rescale(doy), color = "doy")) +
  geom_smooth(aes(x = year, y = rescale(doy), color = "doy"), se = FALSE) +
  # geom_point(aes(x = year, y = rescale(spring_avg_temp), color = "spring_avg_temp")) +
  geom_smooth(aes(x = year, y = rescale(spring_avg_temp), color = "spring_avg_temp"), se = FALSE) +
  # geom_point(aes(x = year, y = rescale(norm), color = "norm")) +
  geom_smooth(aes(x = year, y = rescale(norm), color = "norm"), se = FALSE) +
  # geom_point(aes(x = year, y = rescale(anom), color = "anom")) +
  geom_smooth(aes(x = year, y = rescale(anom), color = "anom"), se = FALSE) +
  ylab("Scaled Value") +
  scale_color_manual(name = "Legend", 
                     values = c("doy" = "blue", "spring_avg_temp" = "red", "norm" = "black", "anom" = "purple"),
                     labels = c("doy", "spring_ave_temp", "norm", "anom"))

#   geom_smooth(method = "lm", aes(x = year, y = rescale(norm), color = "norm"), se = TRUE) +

```

We can see that the sample do cause wrong trend (sample place norm temperature is changing with time). So the data need to be resampled to get the trend

```{r}
library(sampling)

library(dplyr)

# Add a 30-year period column to the dataframe
pheno_with_climate <- pheno_with_climate %>%
  mutate(period = floor((year - 1883) / 30) * 30 + 1890)  # Adjust 1900 based on your data range

summary_decade <- pheno_with_climate %>% 
  group_by(longitude, latitude) %>% 
  summarise(count = n())

pheno_with_climate %>% 
  ggplot() +
  

pheno_with_climate %>%
  group_by(period) %>%
  summarize(mean_norm = mean(norm, na.rm = TRUE)) %>%
  ggplot(aes(x = period, y = mean_norm)) +
  geom_line() +
  labs(title = "Decadal Mean of 'norm' Before Resampling", x = "Decade", y = "Mean Norm")

```

## five revisit individual
```{r}
summary_decade <- pheno_with_climate %>%
    group_by(longitude, latitude) %>%
    summarise(count = n()) %>%
    filter(count > 1) %>% 
    ungroup() %>% 
    mutate(group_id = row_number()) %>% 
    left_join(pheno_with_climate, by = c("longitude", "latitude"))

p1 <- summary_decade %>% 
  ggplot()+
  geom_point(aes(x = spring_avg_temp, y = doy, color = as.factor(group_id))) +
  geom_smooth(method = lm, aes(x = spring_avg_temp, y = doy, color = as.factor(group_id)), se = FALSE) +
  theme(legend.position = "none")

p2 <- summary_decade %>% 
  ggplot()+
  geom_point(aes(x = year, y = doy, color = as.factor(group_id))) +
  geom_smooth(method = lm, aes(x = year, y = doy, color = as.factor(group_id)), se = FALSE) + 
  theme(legend.position = "none")
p1+p2
  
```

## site level revisit with 0.1 degree around 11km: there are some duplicates 
```{r}
summary_decade <- pheno_with_climate %>% 
  mutate(longitude = round(longitude, 1),
         latitude = round(latitude, 1)) %>% 
    group_by(longitude, latitude) %>%
    summarise(count = n()) %>%
    filter(count > 1) %>% 
    ungroup() %>% 
    mutate(group_id = row_number()) %>% 
    left_join(pheno_with_climate%>%
  mutate(longitude = round(longitude, 1),
         latitude = round(latitude, 1)), by = c("longitude", "latitude")) %>% 
    distinct(group_id, doy, spring_avg_temp, year) %>%
    group_by(group_id) %>%
    filter(n() > 1) %>%
    ungroup() 


p1 <- summary_decade %>% 
  ggplot()+
  geom_point(aes(x = spring_avg_temp, y = doy, color = as.factor(group_id))) +
  geom_smooth(method = lm, aes(x = spring_avg_temp, y = doy, color = as.factor(group_id)), se = FALSE) +
  theme(legend.position = "none")

p2 <- summary_decade %>% 
  ggplot()+
  geom_point(aes(x = year, y = doy, color = as.factor(group_id))) +
  geom_smooth(method = lm, aes(x = year, y = doy, color = as.factor(group_id)), se = FALSE) + 
  theme(legend.position = "none")
p1+p2
```

## spatial and temporal sensitivity change with time

```{r}
model2 <- lm(doy~norm+anom+decade*norm,pheno_with_climate)
summary(model2)
confint(model2)
```
# try this method for npn data
```{r}

maple_data <- readRDS("../data/different_species/Acer.rds") %>% 
  filter(common_name == "sugar maple") %>% 
  dplyr::select(year,longitude,latitude,flower) %>% 
  rename(doy = flower)

coordinate <- maple_data %>% 
  distinct(longitude, latitude)


results <- data.frame(year = integer(), longitude = numeric(), latitude = numeric(), spring_avg_temp = numeric())
prism_set_dl_dir("../data/prism")

for (focal_year in 2010:2022) {
  # Load the raster files for March, April, and May
  tmean_rast_yr_mo <- prism_archive_subset(temp_period = "monthly", type = "tmean", years = focal_year, mon = 3:5)
  tmean_rast2_yr_mo <- pd_stack(tmean_rast_yr_mo)
  r_mean <- raster::calc(tmean_rast2_yr_mo, mean) # 1:4 mean for focal_yr
  #raster::plot(r_mean)

  tmean_data <- unlist(raster::extract(x = r_mean,
                                       y = coordinate))

  results <- rbind(results, data.frame(year = focal_year,
                                     longitude = coordinate$longitude,
                                     latitude = coordinate$latitude,
                                     spring_avg_temp = tmean_data))
}


norm_data <- results %>%
  group_by(longitude, latitude) %>%
  summarize(norm = mean(spring_avg_temp, na.rm = TRUE), .groups = 'drop') %>%  # Calculate group means and ungroup
  right_join(results, by = c("longitude", "latitude")) %>%  # Join summarized data back to original
  mutate(anom = spring_avg_temp - norm)  # Calculate anomaly

            

pheno_with_climate <- maple_data %>%
  left_join(norm_data, by = c("longitude", "latitude", "year"))
```


```{r}
# remove doy after 200, and out of us points
pheno_with_climate <- pheno_with_climate %>%
  filter(doy < 200) 

paired_data <- data.frame(
  x_black = pheno_with_climate$spring_avg_temp,
  y_black = pheno_with_climate$doy,
  x_red = pheno_with_climate$norm,
  y_red = pheno_with_climate$doy
)

# Plot the black points
ggplot(paired_data) +
  geom_point(aes(x = x_black, y = y_black, color = "black")) +
  geom_point(aes(x = x_red, y = y_red, color = "red"), alpha = 0.2) +
  geom_segment(aes(x = x_black, y = y_black, xend = x_red, yend = y_red, color = "red"), alpha = 0.2) +
  xlab("spring_ave_temperature") +
  ylab("doy") +
  scale_color_manual(name = "Legend", 
                     values = c("black" = "black", "red" = "red"),
                     labels = c("Observation", "Normality"))
```



```{r}
model2 <- lm(doy~norm+anom,pheno_with_climate)
summary(model2)
confint(model2)

```

