---
title: "figures for outline"
format: html
editor: visual
---

```{r}
library(tidyverse)
```

# Figure 1 SFTS concept

## figure 1a
```{r}
library(ggtext)
figure_1a <- ggplot(data = black) + 
  # geom_point(aes(x = x, y = y)) +
  geom_line(mapping = aes(x = x, y = y), color = "red") +
  theme_classic() + 
  theme(axis.text.x = element_blank(), 
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_line(color = "black")) +
  xlim(-0.3, 1) +
  ylim(-0.1, 1) +
labs(x = expression("Temperature normality" ~ t[norm]), y = expression("Flowering day normality" ~ d[norm])) +
  theme(axis.title.x = element_text(size = 18), axis.title.y = element_text(size = 18)) +
  annotate("text", x = -0.12 , y = 0.02,, label = "Early flowering", size = 6, hjust = 0.7) +
  annotate("text",  x = -0.12 , y = 1, label = "Late flowering", size = 6, hjust = 0.7) +
  annotate("text", x = 0.1 , y = -0.1, label = "Cold place", size = 6, vjust = 0.2) +
  annotate("text", x = 0.9 , y = -0.1, label = "Hot place", size = 6,  vjust = 0.2) 
```


## figure 1b
```{r}
figure_1b <- ggplot(data = black) + 
  # geom_point(aes(x = x, y = y)) +
  geom_line(mapping = aes(x = x, y = y), color = "purple", size = 2) +
  theme_classic() + 
  theme(axis.text.x = element_blank(), 
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_line(color = "black")) +
  xlim(-0.3, 1) +
  ylim(-0.1, 1) +
labs(x = expression("Temperature anomaly" ~ t[anom]), y = expression("Flowering day anomaly" ~ d[anom])) +
  theme(axis.title.x = element_text(size = 18), axis.title.y = element_text(size = 18)) +
  annotate("text", x = -0.12 , y = 0.02,, label = "Early flowering", size = 6, hjust = 0.7) +
  annotate("text",  x = -0.12 , y = 1, label = "Late flowering", size = 6, hjust = 0.7) +
  annotate("text", x = 0.1 , y = -0.1, label = "Cold year", size = 6, vjust = 0.2) +
  annotate("text", x = 0.9 , y = -0.1, label = "Hot year", size = 6,  vjust = 0.2)  +
  # add a horizontal and vertical dash line cross 0,0 and label that they represent 0,0
  geom_point(aes(x = 0.5, y = 0.5), color = "purple", size = 4) +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  geom_vline(xintercept = 0.5, linetype = "dashed") +
  annotate("text", x = 0.55, y = 0.55, label = "(0, 0)", size = 5)
```


## figure 1c

fake data
```{r}
# Combined data frame
black <- data.frame(x = c(0, 1), y = c(1, 0), color = "black")
purple_center <- data.frame(x = seq(0.2, 0.8, by = 0.2), y = 1 - seq(0.2, 0.8, by = 0.2))

# Function to calculate line segments
calculate_segments <- function(center, slope, length) {
  start_x <- center$x - length / 2 * cos(atan(slope))
  start_y <- center$y - length / 2 * sin(atan(slope))
  end_x <- center$x + length / 2 * cos(atan(slope))
  end_y <- center$y + length / 2 * sin(atan(slope))
  
  tibble(start_x = start_x, start_y = start_y, end_x = end_x, end_y = end_y)
}

# Apply the function to all centers
segments_1 <- bind_rows(lapply(1:nrow(purple_center), function(i) 
  calculate_segments(purple_center[i, ], -1.5, 0.18)))
segments_2 <- bind_rows(lapply(1:nrow(purple_center), function(i) 
  calculate_segments(purple_center[i, ], -1, 0.18)))
segments_3 <- bind_rows(lapply(1:nrow(purple_center), function(i) 
  calculate_segments(purple_center[i, ], -0.5, 0.18)))
```

plot
```{r}
# Plot
figure_1c <- ggplot(mapping = aes(x = x, y = y)) + 
  geom_line(data = black, aes(group = 1), color = "red") +
  geom_point(data = purple_center, color = "purple", size = 3) +
  geom_segment(data = segments_2, aes(x = start_x, y = start_y, xend = end_x, yend = end_y), color = "purple", size = 2) +  
  # geom_hline(yintercept = 0.4, linetype = "dashed") +
  # geom_vline(xintercept = 0.6, linetype = "dashed") +
  labs(x = expression(t[observe]), y = expression(d[observe])) + 
  theme_classic() + 
  theme(axis.text.x = element_blank(), 
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_line(color = "black"),
        axis.title.x = element_text(size = 18), axis.title.y = element_text(size = 18))
```

## figure 1 (failed to put them together properly)

```{r}
library(gridExtra)
library(grid)

# Create empty placeholders
empty_grob <- grid::nullGrob()

# Arrange the figures
figure_1 <- gridExtra::grid.arrange(
  figure_1a, figure_1b, 
  empty_grob, figure_1c, empty_grob, 
  nrow = 2
)

figure_1
```

# Figure 2: the by group result from the two datasets
## herbairum
prepare data
```{r}
temperature_data <- read.csv("../data/herb_temperature_data.csv")

temperature_data_model <- temperature_data %>%
  group_by(species) %>%
  # filter(n_distinct(doy) > 10) %>%  # hard to justify
  filter(n_distinct(anom) > 10) %>%  # Use n_distinct() for distinct counting
  filter(n_distinct(norm) > 10) %>%  # Use n_distinct() for distinct counting
  filter(n_distinct(doy, norm, anom) > 30) %>%  # Use n_distinct() for distinct counting
  ungroup() 

sample_size <- temperature_data_model %>%
  group_by(species) %>%
  summarise(n = n()) %>%
  ungroup() %>%
  rename(species = species)

summary_results <- read.csv("../data/species_summary_herb.csv")

summary_results_wtaxa <- temperature_data_model %>%
  distinct(species, genus, family) %>% 
  right_join(summary_results, by = "species") %>% 
  mutate(taxa = genus) %>%
  left_join(sample_size, by = "species") 

summary_results_wtaxa_combined <- summary_results_wtaxa %>%
  mutate(sensitivity_type = "Spatial", estimate_combined = norm_estimate, conf_low_combined = norm_conf_low, conf_high_combined = norm_conf_high) %>%
  bind_rows(
    summary_results_wtaxa %>%
      mutate(sensitivity_type = "Temporal", estimate_combined = anom_estimate, conf_low_combined = anom_conf_low, conf_high_combined = anom_conf_high)
  ) %>% 
  arrange(taxa, species) %>%
  mutate(species_name = forcats::fct_reorder(species, desc(species))) # Reorder species alphabetically within taxa

```

plot
```{r fig.width=15, fig.height=6}
herb_byspecies_plot <- summary_results_wtaxa_combined %>%
  ggplot(aes(x = species_name, y = estimate_combined, color = sensitivity_type)) +  
  # define the color
  scale_color_manual(values = c("Spatial" = "red", "Temporal" = "purple")) +
  
  # Points with dodging for groups
  geom_point(size = 2, position = position_dodge(width = 0.5), shape = 3) +
  
  # Error bars with the same dodging as points
  geom_errorbar(aes(ymin = conf_low_combined, ymax = conf_high_combined), 
                width = 0, position = position_dodge(width = 0.5)) +
  
  geom_text(aes(y = -21, label = ifelse(equal == 0, "*", "")), vjust = 0.75, color = "black") +
  
  # Add horizontal line at sensitivity = 0
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  
  # Facet by taxa to group species
  facet_grid(. ~ taxa, scales = "free_x", space = "free_x") + 

  # Clip the plot to specific y-limits
  coord_cartesian(ylim = c(-20, 10)) +
  
  # Set axis labels, title, and color legend
  labs(
    title = "", 
    x = "Species", 
    y = "Sensitivity (days/째C)", 
    color = "Sensitivity type"
  ) +
  
  # Apply themes for classic look, italics, and markdown legend
  theme_classic() +
  theme(
    axis.text.x = element_text(face = "italic", angle = 90, hjust = 1, vjust = 0.5), 
    strip.text = element_text(face = "italic"),  # Italicize facet panel labels
    legend.text = element_markdown(),
    legend.position = "top"
  )

```

## NPN
prepare data
```{r}
temperature_data <- read.csv("../data/npn_temperature_data.csv") 

temperature_data_model <- temperature_data %>%
  group_by(species) %>%
  # filter(n_distinct(doy) > 10) %>%  # hard to justify
  filter(n_distinct(anom) > 10) %>%  # Use n_distinct() for distinct counting
  filter(n_distinct(norm) > 10) %>%  # Use n_distinct() for distinct counting
  filter(n_distinct(doy, norm, anom) > 30) %>%  # Use n_distinct() for distinct counting
  ungroup() 

sample_size <- temperature_data_model %>%
  group_by(species) %>%
  summarise(n = n()) %>%
  ungroup() %>%
  rename(species = species)

summary_results <- read.csv("../data/species_summary_npn.csv")

summary_results_wtaxa <- temperature_data_model %>%
  distinct(species, genus, taxa) %>% 
  right_join(summary_results, by = "species") %>% 
  left_join(sample_size, by = "species") 

summary_results_wtaxa_combined <- summary_results_wtaxa %>%
  mutate(sensitivity_type = "Spatial", estimate_combined = norm_estimate, conf_low_combined = norm_conf_low, conf_high_combined = norm_conf_high) %>%
  bind_rows(
    summary_results_wtaxa %>%
      mutate(sensitivity_type = "Temporal", estimate_combined = anom_estimate, conf_low_combined = anom_conf_low, conf_high_combined = anom_conf_high)
  ) %>% 
  arrange(taxa, species) %>%
  mutate(species_name = forcats::fct_reorder(species, desc(species))) # Reorder species alphabetically within taxa

```

plot
```{r fig.width=15, fig.height=6}
npn_byspecies_plot <- summary_results_wtaxa_combined %>%
  ggplot(aes(x = species_name, y = estimate_combined, color = sensitivity_type)) +  
  # define the color
  scale_color_manual(values = c("Spatial" = "red", "Temporal" = "purple")) +
  
  # Points with dodging for groups
  geom_point(size = 2, position = position_dodge(width = 0.5), shape = 3) +
  
  # Error bars with the same dodging as points
  geom_errorbar(aes(ymin = conf_low_combined, ymax = conf_high_combined), 
                width = 0, position = position_dodge(width = 0.5)) +
  
  geom_text(aes(y = -21, label = ifelse(equal == 0, "*", "")), vjust = 0.75, color = "black") +
  
  # Add horizontal line at sensitivity = 0
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  
  # Facet by taxa to group species
  facet_grid(. ~ taxa, scales = "free_x", space = "free_x") + 

  # Clip the plot to specific y-limits
  coord_cartesian(ylim = c(-20, 10)) +
  
  # Set axis labels, title, and color legend
  labs(
    title = "", 
    x = "Species", 
    y = "Sensitivity (days/째C)", 
    color = "Sensitivity type"
  ) +
  
  # Apply themes for classic look, italics, and markdown legend
  theme_classic() +
  theme(
    axis.text.x = element_text(face = "italic", angle = 90, hjust = 1, vjust = 0.5), 
    strip.text = element_text(face = "italic"),  # Italicize facet panel labels
    legend.text = element_markdown(),
    legend.position = "top"
  )

```

## combine
```{r fig.width=15, fig.height=12}
gridExtra::grid.arrange(herb_byspecies_plot, npn_byspecies_plot, nrow = 2)

```
# Figure 3: the HMM and PMM results
## HMM
```{r}
library(ape)
library(ggtree)
library(phytools)
library(rstan)
library(ggplot2)
library(dplyr)
library(ggtext)
HMM <- readRDS("../data/phylogenetics/fit_model_HMM_all.rds")
# summarize the results
HMM_summary <- summary(HMM) 
  
# Extract the summary information into a data frame
HMM_group <- as.data.frame(HMM_summary$summary)

# get genus information
taxa_info <- read.csv("../data/phylogenetics/total_species.csv")

# get the phylo_species_id
tree <- read.tree("../data/phylogenetics/PhyloMaker_tree_scenario1_total.nwk") 
tree_species <- rownames(vcv(tree, corr = TRUE)) %>% 
  gsub("_", " ", .) 

phylo_species_id <- data.frame(species = tree_species, sppid = seq_along(tree_species))


diff_sensitivity_HMM <- HMM_group[grepl("b_chill", rownames(HMM_group)), ] %>% 
  dplyr::select(mean, `2.5%`, `97.5%`) %>%
  dplyr::rename(mean_diff = mean, lower_diff = `2.5%`, upper_diff = `97.5%`) %>% 
  cbind(phylo_species_id) %>% 
  left_join(taxa_info, by = "species")
```



```{r fig.width=9, fig.height=32}
# Extract the posterior samples for b_zc from the Stan model
posterior_b_zc <- extract(HMM)$b_zc
posterior_b_zc <- as.data.frame(posterior_b_zc)

fitsum_h <- summary(HMM, pars = c("a_z", "sigma_interceptsa", 
                                       "b_zf", "sigma_interceptsbf", 
                                       "b_zc", "sigma_interceptsbc"))
fitsumdf_h <- as.data.frame(fitsum_h)


# Get the 95% confidence interval from the summary data frame
posterior_b_zc_lower <- fitsumdf_h %>% filter(row.names(.) == "b_zc") %>% pull(`summary.2.5.`)
posterior_b_zc_upper <- fitsumdf_h %>% filter(row.names(.) == "b_zc") %>% pull(`summary.97.5.`)

# Plot posterior distribution with 95% CI
HMM_den <- ggplot(posterior_b_zc, aes(x = posterior_b_zc)) +
  geom_density(fill = "skyblue", color = "skyblue") +  # Posterior density plot
  geom_vline(xintercept = posterior_b_zc_lower, linetype = "dashed", color = "red", size = 1) +  # Lower bound of 95% CI
  geom_vline(xintercept = posterior_b_zc_upper, linetype = "dashed", color = "red", size = 1) +  # Upper bound of 95% CI
  xlim(-8, 8) +
  labs(title = " ",
       x = "",
       y = "") +
    theme_minimal() +
  theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        axis.title.x = element_blank(), axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        panel.grid = element_blank())


HMM_species <- diff_sensitivity_HMM %>% 
  ggplot(aes(x = mean_diff, y = species)) +  # Swapped x and y
  geom_point(shape = 3, size = 2, position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(xmin = lower_diff, xmax = upper_diff)) +  # Changed to xmin and xmax
  geom_vline(xintercept = 0, linetype = "dashed") +  # Changed to vline
  labs(
    title = "",
    y = "Species",  # Swapped x and y labels
    x = "Difference (Temporal - Spatial, days/째C)",
    color = "Dataset"
  ) +
  xlim(-8, 8) +
  facet_grid(taxa ~ ., scales = "free_y", space = "free_y") +  # Changed to facet by rows
  theme_classic() +
  theme(
    axis.text.y = element_text(face = "italic"),  # Changed to y-axis text
    legend.text = element_markdown()
  )


# put these two figures together
library(patchwork)
HMM_combined <- HMM_den / plot_spacer() / HMM_species + plot_layout(heights = c(1, 0, 30))
```

PMM

## PMM
```{r}
PMM <- readRDS("../data/phylogenetics/fit_model_PMM_all.rds")
# summarize the results
PMM_summary <- summary(PMM)

# Extract the summary information into a data frame
PMM_group <- as.data.frame(PMM_summary$summary)

diff_sensitivity_PMM <- PMM_group[grepl("b_chill", rownames(PMM_group)), ] %>% 
  dplyr::select(mean, `2.5%`, `97.5%`) %>%
  dplyr::rename(mean_diff = mean, lower_diff = `2.5%`, upper_diff = `97.5%`) %>% 
  cbind(phylo_species_id) %>% 
  left_join(taxa_info, by = "species")
```

```{r fig.width=9, fig.height=32}
# Extract the posterior samples for b_zc from the Stan model
posterior_b_zc <- extract(PMM)$b_zc
posterior_b_zc <- as.data.frame(posterior_b_zc)

fitsum_h <- summary(PMM, pars = c("a_z", "sigma_interceptsa", 
                                       "b_zf", "sigma_interceptsbf", 
                                       "b_zc", "sigma_interceptsbc"))
fitsumdf_h <- as.data.frame(fitsum_h)


# Get the 95% confidence interval from the summary data frame
posterior_b_zc_lower <- fitsumdf_h %>% filter(row.names(.) == "b_zc") %>% pull(`summary.2.5.`)
posterior_b_zc_upper <- fitsumdf_h %>% filter(row.names(.) == "b_zc") %>% pull(`summary.97.5.`)

# Plot posterior distribution with 95% CI
PMM_den <- ggplot(posterior_b_zc, aes(x = posterior_b_zc)) +
  geom_density(fill = "skyblue", color = "skyblue") +  # Posterior density plot
  geom_vline(xintercept = posterior_b_zc_lower, linetype = "dashed", color = "red", size = 1) +  # Lower bound of 95% CI
  geom_vline(xintercept = posterior_b_zc_upper, linetype = "dashed", color = "red", size = 1) +  # Upper bound of 95% CI
  xlim(-8, 8) +
  labs(title = " ",
       x = "",
       y = "") +
    theme_minimal() +
  theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        axis.title.x = element_blank(), axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        panel.grid = element_blank())


PMM_species <- diff_sensitivity_PMM %>% 
  ggplot(aes(x = mean_diff, y = species)) +  # Swapped x and y
  geom_point(shape = 3, size = 2, position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(xmin = lower_diff, xmax = upper_diff)) +  # Changed to xmin and xmax
  geom_vline(xintercept = 0, linetype = "dashed") +  # Changed to vline
  labs(
    title = "",
    y = "Species",  # Swapped x and y labels
    x = "Difference (Temporal - Spatial, days/째C)",
    color = "Dataset"
  ) +
  xlim(-8, 8) +
  facet_grid(taxa ~ ., scales = "free_y", space = "free_y") +  # Changed to facet by rows
  theme_classic() +
  theme(
    axis.text.y = element_text(face = "italic"),  # Changed to y-axis text
    legend.text = element_markdown()
  )


```

## combined figure
```{r fig.width=9, fig.height=32}
library(patchwork)

HMM_combined | PMM_combined
```


# Figure 4: Estimated sensitivity for wind pollinated trees flower time with their phylogenetic signal

