---
title: "Revision"
format: html
editor: visual
---

# Check the georeferencing infomation

Download the GBIF archive data from https://www.gbif.org/occurrence/download/0011964-240506114902167

```{r}
library(data.table)
library(dplyr)

occ <- fread("~/Downloads/0011964-240506114902167/occurrence.txt", sep = "\t", quote = "")
```

```{r}
grep("georef|coordinate|latitude|longitude",
     names(occ),
     ignore.case = TRUE,
     value = TRUE)
```

## Check with PRISM resolution

```{r}
occ %>%
  summarize(
    `<=4 km` = sum(coordinateUncertaintyInMeters <= 4000, na.rm = TRUE),
    `>4 km` = sum(coordinateUncertaintyInMeters > 4000, na.rm = TRUE),
    missing = sum(is.na(coordinateUncertaintyInMeters))
  )
```

```{r}
head(sort(table(occ$georeferenceProtocol), decreasing = TRUE))
```

```{r}
head(sort(table(occ$georeferenceVerificationStatus), decreasing = TRUE))
```

# Check the species list for dioecious

```{r}
herb_data <- read.csv("~/Downloads/to_release/species_summary.csv") %>% 
  filter(dataset == "herb")
```

```{r}
print(herb_data$species)
```

# Check spcies coverage for Ramirez-Parada et al

## Run their "1. Subsetting Dataset.Rmd"

Get this data: **Export dataset for 300+ species**

```{r}
fwrite(dat_master, file = "../specimen data - 1,605 species, 300+ specimens each.csv")
```

## Test number of overlaped species

Genus: Acer Betula Fraxinus Morus Populus Quercus Ulmus

```{r}
library(dplyr)
wind_trees <- dat_master %>%
  filter(genus %in% c(
    "Acer",
    "Betula",
    "Fraxinus",
    "Morus",
    "Populus",
    "Quercus",
    "Ulmus"
  )) %>%
  distinct(genus, Accepted_species)
```


```{r}
species_composition <- read_csv("/Volumes/seas-zhukai/proj-phenology-sensitivity/phenology-Replicate-Ramirez-Parada-paper/species_composition.csv")
```

# Visual check machine learning result:

```{r}
# write.csv(joint_data_flower, .path$herb_flower) # read this file
joint_data_flower <- read_csv(.path$herb_flower)
```


```{r}
library(dplyr)
library(readr)
library(stringr)

## Read multimedia file
multimedia <- read_tsv(
  "~/Downloads/0011964-240506114902167/multimedia.txt",
  show_col_types = FALSE
)

## Genera represented in the flowering dataset
flower_genera <- unique(joint_data_flower$genus)

############################################################
# Flower candidate pool (ALL flowering specimens)
############################################################

flower_pool <- joint_data_flower %>%
  mutate(type = "flower")

############################################################
# Non-flower candidate pool (ALL non-flowering specimens)
############################################################

nonflower_pool <- raw %>%
  anti_join(
    joint_data_flower %>% select(filename_image),
    by = "filename_image"
  ) %>%
  filter(genus %in% flower_genera) %>%
  mutate(type = "non_flower")

############################################################
# Combine
############################################################

image_pool <- bind_rows(
  flower_pool,
  nonflower_pool
)

############################################################
# Extract GBIF ID from filename
############################################################

image_pool <- image_pool %>%
  mutate(
    gbifID = str_extract(filename_image, "(?<=_)[0-9]+(?=_)")
  )

############################################################
# Join image URLs
############################################################

download_list <- image_pool %>%
  mutate(gbifID = as.character(gbifID)) %>%
  left_join(
    multimedia %>%
      mutate(gbifID = as.character(gbifID)) %>%
      select(gbifID, identifier) %>%
      distinct(gbifID, .keep_all = TRUE),
    by = "gbifID"
  ) %>%
  filter(!is.na(identifier)) %>%
  select(
    gbifID,
    identifier,
    genus,
    species,
    type,
    filename_image,
    everything()
  )

############################################################
# Save for Shiny
############################################################



saveRDS(
  download_list,
  "~/Downloads/FlowerValidation/data/download_list.rds"
)

cat(nrow(download_list), "candidate specimens saved.\n")
```


```{r}


setwd("~/Downloads/FlowerValidation_updated")

shiny::runApp()
```



# Species count

## spring
```{r}
spring_species <- read.csv("/Volumes/seas-zhukai/proj-phenology-sensitivity/review_np/spring/species_summary.csv")

joint_data_flower %>%
  semi_join(spring_species, by = "species") %>%
  count(species, name = "count")
```

```{r}
library(ggplot2)

count_df <- joint_data_flower %>%
  semi_join(spring_species, by = "species") %>%
  count(species, name = "count")

n_above300 <- sum(count_df$count > 300)

ggplot(count_df, aes(x = count)) +
  geom_histogram(binwidth = 5, color = "white", fill = "steelblue") +
  geom_vline(xintercept = 300, linetype = "dashed", color = "red", linewidth = 1) +
  annotate(
    "text",
    x = 300,
    y = Inf,
    label = paste0(n_above300, " species > 300"),
    vjust = 1.5,
    hjust = -0.1,
    color = "red",
    fontface = "bold"
  ) +
  labs(
    x = "Number of flowering specimens",
    y = "Number of species",
    title = "Distribution of specimen counts per species"
  ) +
  theme_bw()
```

```{r fig.width=10, fig.height=10}
library(ggtext)

count_df <- joint_data_flower %>%
  semi_join(spring_species, by = "species") %>%
  count(species, name = "count") %>%
  arrange(count) %>%
  mutate(species = factor(species, levels = species))

ggplot(count_df, aes(x = species, y = count)) +
  geom_col() +
  coord_flip() +
  labs(
    x = "Species",
    y = "Number of specimens"
  ) +
  theme_classic() +
  theme(
    axis.text.y = element_markdown(face = "italic")
  )
```



## winter

```{r}
winter_species <- read.csv("/Volumes/seas-zhukai/proj-phenology-sensitivity/review_np/winter/species_summary.csv")

joint_data_flower %>%
  semi_join(winter_species, by = "species") %>%
  count(species, name = "count") %>% 
  arrange(desc(count))
```

```{r}
count_df <- joint_data_flower %>%
  semi_join(spring_species, by = "species") %>%
  count(species, name = "count")

n_above300 <- sum(count_df$count > 300)

ggplot(count_df, aes(x = count)) +
  geom_histogram(binwidth = 5, color = "white", fill = "steelblue") +
  geom_vline(xintercept = 300, linetype = "dashed", color = "red", linewidth = 1) +
  annotate(
    "text",
    x = 300,
    y = Inf,
    label = paste0(n_above300, " species > 300"),
    vjust = 1.5,
    hjust = -0.1,
    color = "red",
    fontface = "bold"
  ) +
  labs(
    x = "Number of flowering specimens",
    y = "Number of species",
    title = "Distribution of specimen counts per species"
  ) +
  theme_bw()
```

```{r}
load("/Volumes/seas-zhukai/proj-phenology-sensitivity/phenology-Replicate-Ramirez-Parada-paper/Correlation - tmean_norm vs tmean_WL_max - 1605 spp.Rdata")
```

