---
title: "Revision"
format: html
editor: visual
---

# Check the georeferencing infomation

Download the GBIF archive data from https://www.gbif.org/occurrence/download/0011964-240506114902167

```{r}
library(data.table)
library(dplyr)

occ <- fread("~/Downloads/0011964-240506114902167/occurrence.txt", sep = "\t", quote = "")
```

```{r}
grep("georef|coordinate|latitude|longitude",
     names(occ),
     ignore.case = TRUE,
     value = TRUE)
```

## Check with PRISM resolution

```{r}
occ %>%
  summarize(
    `<=4 km` = sum(coordinateUncertaintyInMeters <= 4000, na.rm = TRUE),
    `>4 km` = sum(coordinateUncertaintyInMeters > 4000, na.rm = TRUE),
    missing = sum(is.na(coordinateUncertaintyInMeters))
  )
```

```{r}
head(sort(table(occ$georeferenceProtocol), decreasing = TRUE))
```

```{r}
head(sort(table(occ$georeferenceVerificationStatus), decreasing = TRUE))
```

# Check the species list for dioecious

```{r}
herb_data <- read.csv("~/Downloads/to_release/species_summary.csv") %>% 
  filter(dataset == "herb")
```

```{r}
print(herb_data$species)
```

# Check spcies coverage for Ramirez-Parada et al

## Run their "1. Subsetting Dataset.Rmd"

Get this data: **Export dataset for 300+ species**

```{r}
fwrite(dat_master, file = "../specimen data - 1,605 species, 300+ specimens each.csv")
```

## Test number of overlaped species

Genus: Acer Betula Fraxinus Morus Populus Quercus Ulmus

```{r}
library(dplyr)
wind_trees <- dat_master %>%
  filter(genus %in% c(
    "Acer",
    "Betula",
    "Fraxinus",
    "Morus",
    "Populus",
    "Quercus",
    "Ulmus"
  )) %>%
  distinct(genus, Accepted_species)
```

# visual check machine learning result:




```{r}
occurrence <- read.delim(
  "~/Downloads/0011964-240506114902167/occurrence.txt",
  sep = "\t",
  stringsAsFactors = FALSE
)
```

```{r}
names(occurrence)
```

```{r}
library(stringr)
image_sample <- bind_rows(flower_sample, nonflower_sample)

image_sample_2 <- image_sample %>%
  mutate(
    gbifID = str_extract(filename_image, "(?<=_)[0-9]+(?=_)")
  )
```



```{r}
library(readr)

multimedia <- read_tsv(
  "~/Downloads/0011964-240506114902167/multimedia.txt",
  show_col_types = FALSE
)


download_list <- image_sample_2 %>%
  mutate(gbifID = as.character(gbifID)) %>%
  left_join(
    multimedia %>%
      mutate(gbifID = as.character(gbifID)) %>%
      select(gbifID, identifier) %>%
      distinct(gbifID, .keep_all = TRUE),
    by = "gbifID"
  ) %>%
  filter(!is.na(identifier))
```


```{r}
library(dplyr)
library(fs)
library(furrr)

# Create folders
dir_create("images/flower")
dir_create("images/non_flower")

# Download function
download_image <- function(url, gbifID, type) {
  
  # skip missing URLs
  if (is.na(url) || url == "") return(NULL)
  
  # get extension
  ext <- tools::file_ext(sub("\\?.*$", "", url))
  if (ext == "") ext <- "jpg"
  
  # output path
  outfile <- file.path(
    "images",
    type,
    paste0(gbifID, ".", ext)
  )
  
  # skip existing files
  if (file.exists(outfile)) return(NULL)
  
  tryCatch(
    {
      download.file(
        url,
        outfile,
        mode = "wb",
        quiet = TRUE
      )
    },
    error = function(e) {
      message("Failed: ", gbifID)
    }
  )
}


# Use multiple cores
plan(multisession, workers = 8)

future_pwalk(
  download_list %>%
    select(url = identifier, gbifID, type),
  download_image
)
```

